Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-05-23 paper/main.tex
start 3.2

We next place a zero-mean Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel.  
\textcolor{red}{(@Steve - is a zero mean the right thing?  If not, what is the mean?)}
Standard practice would be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}|S,W &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S|W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernels,
\begin{align}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2, \;\;\textrm{and} \\
&\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
  \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2.
\end{align}
On the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the 
correlation scale and lengthscale respectively.  On the $W$ layer, we allow for a
unique lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.

We have found that setting the prior mean of $W$ to $X$ instead of the $\mathbf{0}$
in Eq.~\ref{eq:dgphm},
\citep[which would indicate stationarity apriori,][]{schmidt2003bayesian} does not 
have a substantial impact on the model fit. We construct this model in the 
compositional form, and also use Markov Chain Monte Carlo (MCMC) to estimate and 
obtain full uncertainty for each of the model parameters \citep{sauer2023active}.
\textcolor{red}{(Need more details on inference, in next section...)}

@@

% $$
% \mathbf{\Sigma_\varepsilon} =
% \begin{bmatrix}
% \Lambda_p^{-1} & \mathbf{0} & \mathbf{0} \\
% \mathbf{0} & \mathbf{\Sigma}_{\ell} & \mathbf{0} \\
% \mathbf{0} & \mathbf{0} & \Lambda_h^{-1}
% \end{bmatrix}
% $$

\subsection{Model Inference}

>>>>

We next place a zero-mean Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel.  
\textcolor{red}{(@Steve - is a zero mean the right thing?  If not, what is the mean?)}
Standard practice would be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}|S &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S|W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernels,
\[
\begin{aligned}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2, \;\;\textrm{and} \\
&\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
  \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2.
\end{aligned}
\]
On the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the 
correlation scale and lengthscale respectively.  On the $W$ layer, we allow for a
unique lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.
We considered setting the prior mean of $W$ to $X$ instead of $\mathbf{0}$
\citep[which would indicate stationarity apriori;][]{schmidt2003bayesian}, but found
it had a neglible effect on the model fits.

\subsection{Bayesian Inference}

@@
